\ttitle

### 2015-08-26 Chapters/CPET_Jaundice.tex
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes

\subsection{Aim}
The results presented in Chapter \ref{ch_cpet_outcomes} demonstrated that poor performance at cardiopulmonary exercise testing (CPET) was associated with adverse outcomes after pancreaticoduodenectomy resulting in an increased incidence of POPF and prolonged hospital stay. 
Obstructive jaundice was present in a large proportion of these patients.
The aim of the present study was to evaluate the relationship between cardiopulmonary exercise testing, obstructive jaundice (especially severe obstructive jaundice with serum bilirubin $>$ 250 $\mu$mol/l) and preoperative pathophysiology in patients undergoing major pancreatic surgery.

@@

\subsection{Patients}
Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and April 2013 and had undergone cardiopulmonary exercise testing as part of their preoperative work-up at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study. 
Established criteria for resectability in patients with malignant disease were used as outlined in Section \ref{sec:resectability_criteria}. 
Segmental or wedge resection of the portal vein or superior mesenteric vein was carried out if the lesion was otherwise resectable.

\subsection{Preoperative data}
Patient demographics, preoperative clinico-pathological characteristics including cardiorespiratory comorbidity, results of preoperative blood tests, chest x-ray, ECG and cardiopulmonary exercise tests were collected from prospectively held databases. 
The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease, respiratory disease, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) and was used as an objective score of comorbidity. 
Cardiovascular comorbidity was defined as a score of 2 or more for either the cardiac disease or ECG component of the POSSUM score. 
Respiratory comorbidity was defined as a score or 2 or more for the respiratory disease component of the POSSUM score.

\subsection{Obstructive jaundice}
Serum bilirubin levels and liver function tests were measured in all patients on the day before surgery. 
Obstructive jaundice was defined as serum bilirubin $>$ 35 $\mu$mol/litre and severe obstructive jaundice was defined as serum bilirubin $>$ 250 $\mu$mol/litre. 
These thresholds were chosen as the recent randomised controlled trial that demonstrated increased complications in patients who underwent PBD excluded patients with serum bilirubin $>$ 250 $\mu$mol/litre from the study \parencite{van_der_gaag_preoperative_2010}.
The present study aimed to evaluate preoperative pathophysiology in this particular group of patients with `severe obstructive jaundice'.
Due to referral practices at the time of this study, a proportion of patients underwent preoperative biliary interventions and drainage before referral to the West of Scotland Pancreatic Unit. 
Data regarding biliary intervention and stenting was collected where feasible.

\subsection{Cardiopulmonary exercise test}
Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
All patients underwent standard pulmonary function tests and spirometry prior to cardiopulmonary exercise testing. 
A cycle ergometer was used to perform a symptom-limited, incremental work-load test preceded by a 3-minute rest period. 
The test was stopped when patients achieved their maximum exercise tolerance, when significant ischaemic changes occurred on ECG or for other safety reasons. 
Peak oxygen consumption achieved at this stage was defined as $\dot{V}_{O_2}$Peak. 
The $\dot{V}_{O_2}$AT was calculated using the V-slope \parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents \parencite{society_ats/accp_2003} methods.
$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us (Chapter \ref{ch_cpet_outcomes}) as well as Ausania and co-workers \parencite{ausania_effects_2012} which has shown increased incidence of complications in patients with $\dot{V}_{O_2}$AT below this threshold. 
Oxygen consumption at peak exercise ($\dot{V}_{O_2}$Peak) was dichotomised using a cut-off of 16 ml/kg/min. 
Cardiopulmonary exercise testing methodology and calculation of $\dot{V}_{O_2}$AT is described in detail in Sections \ref{sec:cpx_method} and \ref{sec:cpx_at_method}.
A more detailed description of cardiopulmonary exercise testing parameters described in this study are published elsewhere \parencite{balady_clinicians_2010}.

\subsection{Statistics}
Grouping of the variables was carried out using standard or previously published thresholds. 
In the absence of such thresholds, the variables were treated as continuous variables. 
Non-parametric tests were used to analyse the association between categorical and continuous variables while Chi-square tests were used to analyse the association between categorical variables. 
Univariate and multivariate binary logistic regression analysis was used to study the relationship between preoperative patient characteristics and $\dot{V}_{O_2}$AT / $\dot{V}_{O_2}$Peak. 
SPSS software (Version 17.0; SPSS Inc., Chicago, IL, USA) was used to perform statistical analysis.

@@

The baseline demographic and clinical characteristics of non-jaundiced and jaundiced patients are shown in Table \ref{table:cpet_oj_patient}. 
A greater proportion of jaundiced patients were females compared to the non-jaundiced cohort (p=0.028) and smokers (p=0.038). 
Elevated POSSUM Physiology Score (p=0.004) and malignancy (p$<$0.001) were significantly associated with the presence of jaundice. 
However, there was no statistically significant difference in age, body mass index, cardiovascular comorbidity, respiratory comorbidity or PBD between the non-jaundiced and jaundiced patients.

@@

\subsection{Obstructive jaundice vs. pulmonary function tests}
Pulmonary function tests including forced vital capacity (FVC), forced expiratory volume in one second (FEV1) and their derived parameters (predicted FEV1, FEV1/FVC, predicted FEV1/FVC) were compared between jaundiced and non-jaundiced patients. 
There was a trend towards lower forced vital capacity with increasing severity of jaundice but this did not reach statistical significance (p=0.092). 
There was no association between the other pulmonary function tests and jaundice. (Table \ref{table:cpet_oj_pft}).

@@

\subsection{Obstructive jaundice vs. CPET}
Cardiopulmonary exercise test parameters measured at the anaerobic threshold and at peak exercise were compared between jaundiced and non-jaundiced patients using the Kruskal-Wallis test. 
Obstructive jaundice was associated with lower tidal volume (p=0.017), lower absolute $\dot{V}_{O_2}$ (p=0.003), lower corrected $\dot{V}_{O_2}$ (p=0.029), lower oxygen pulse (p=0.037) and lower respiratory rate (p=0.022). 
At peak exercise, jaundice was associated with lower tidal volume (p=0.028), lower absolute $\dot{V}_{O_2}$Peak (p=0.007), lower absolute $\dot{V}_{CO_2}$ (p=0.016) and lower end-tidal $O_2$ or $PET_{O_2}$ (p=0.026). 
However, as shown in Tables \ref{table:cpet_oj_anaerobic} and \ref{table:cpet_oj_peak}, several of these associations were not linear with a trend towards poorer results between non-jaundiced and mildly jaundiced patients and better values in the severely jaundiced cohort.

@@

The results of the present study demonstrate that obstructive jaundice, even when severe, is not an independent risk factor for poor preoperative cardiopulmonary exercise physiology in patients undergoing pancreaticoduodenectomy. 
The results also demonstrate that obstructive jaundice is associated with elevated preoperative systemic inflammation with the severity of systemic inflammation related to the severity of the jaundice.

%History of CPET
The use of CPET in preoperative risk prediction was first made popular over two decades ago by Older and co-workers \parencite{older_preoperative_1993}. 
Since then cardiopulmonary exercise testing has been reported to be useful in identifying high risk patients prior to major general \parencite{snowden_submaximal_2010}, pancreatic \parencite{ausania_effects_2012},[Chapter \ref{ch_cpet_outcomes}], oesophagogastric \parencite{nagamatsu_preoperative_2001} as well as vascular \parencite{carlisle_mid-term_2007} surgery. 
Cardiopulmonary exercise testing is superior to conventional measures of fitness due to the dynamic nature of the test that evaluates the adequacy of oxygen delivery to tissues under physiological stress. 
However, the factors responsible for poor aerobic capacity in surgical patients have not been adequately studied. 
The present study aimed to address this in patients undergoing major pancreatic surgery.

@@

Historically, obstructive jaundice has also been reported to be associated with adverse haemodynamic events in patients undergoing major surgery. 
Intra-operative blood loss, postoperative hypotension, increased susceptibility to shock and renal dysfunction were all more common in patients with obstructive jaundice \parencite{dixon_factors_1983, pain_perioperative_1985, green_systemic_1995}.
A recent review noted that bile acids have a complex range of receptor mediated effects on the cardiovascular system \parencite{khurana_bile_2011}. 
Moreover, some of these effects appear to be partly reversible by biliary drainage as demonstrated by Padillo and co-workers \parencite{padillo_improved_2001}.
This increased incidence of complications resulted in routine pre-operative biliary drainage being recommended in jaundiced patients before undertaking major surgery. 
Pancreaticoduodenectomy was described by Whipple initially as a two-stage operation, with the first stage involving a cholecysto-gastrostomy aimed at relieving biliary obstruction before undertaking the resection at a later second operation \parencite{whipple_treatment_1935}.

%Reasons for not doing PBD
However, several recent studies have reported that PBD is associated with increased incidence of complications and that surgery in the jaundiced patients is safe.
Pitt and co-workers in a prospective randomised trial comparing outcomes in jaundiced patients undergoing surgery with or without PBD reported that PBD was associated with increased cost without any decrease in postoperative complications \parencite{pitt_does_1985}. 
A recent meta-analysis \parencite{sewnath_meta-analysis_2002} pooled data from 5 randomised controlled trials comparing surgery with PBD versus surgery without drainage.
The authors concluded that PBD not only did not improve postoperative complication rates or mortality but resulted in a higher overall complication rate due to the morbidity associated with the procedure itself. 
These findings have been reproduced in a more recent multicenter, randomised trial \parencite{van_der_gaag_preoperative_2010}.
A recent Cochrane Collaboration review of six trials including 520 patients concluded that PBD may be associated with serious adverse events and must not be performed routinely outside trial settings \parencite{wang_preoperative_2008}.

%CPET vs Jaundice
In the present study, obstructive jaundice was not independently related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak.
There was only a weak negative correlation between serum bilirubin and $\dot{V}_{O_2}$AT in our cohort.
These findings are similar to those of Parker and co-workers who reported that there was only a weak negative correlation between serum bilirubin and  $\dot{V}_{O_2}$Peak (Pearson's $\rho$ -0.21, p=0.02) and no correlation with $\dot{V}_{O_2}$AT ($\rho$ = -0.15, p=NS) in patients undergoing pancreaticoduodenectomy \parencite{parker_serum_2014}. 
Junejo and co-workers studied oxygen extraction in 9 jaundiced patients during cardiopulmonary exercise test by placing catheters within the femoral vein. 
They reported that oxygen extraction was normal during cardiopulmonary exercise even in the presence of jaundice \parencite{junejo_peripheral_2014}

@@

%Obstructive jaundice and systemic inflammation
However, jaundice was associated with elevated preoperative systemic inflammation as measured by C-reactive protein levels and serum albumin. 
Obstructive jaundice is known to be associated with wide ranging abnormalities of both humoral and innate immune systems \parencite{nehez_compromise_2002, padillo_cytokines_2001, scott-conner_pathophysiology_1994}. 
It is unclear if this immune dysfunction before surgery is associated with an abnormal postoperative systemic inflammatory response and this is explored further in Chapter \ref{ch_pre_post_sirs}.
However, the immunological abnormalities associated with obstructive jaundice do not appear to improve after biliary drainage \parencite{kimmings_endotoxin_2000}.

@@

%\subsection{Conclusion}
Obstructive jaundice, including severe obstructive jaundice did not affect preoperative cardiopulmonary exercise physiology. 
Therefore, preoperative biliary drainage cannot be expected to improve aerobic capacity in patients undergoing pancreaticoduodenectomy.
Both obstructive jaundice and low $\dot{V}_{O_2}$AT were associated with elevated preoperative systemic inflammation and provide further evidence of the role of jaundice and impaired aerobic capacity in immune dysregulation in these patients.

>>>>

\subsection{Aim}
The results presented in Chapter \ref{ch_cpet_outcomes} demonstrated that poor performance at cardiopulmonary exercise testing (CPET) was associated with adverse outcomes after pancreaticoduodenectomy with an increased incidence of POPF and prolonged hospital stay. 
Obstructive jaundice was present in a large proportion of these patients but not associated with prolonged hospitalisation (Table \ref{table:cpet_outcomes_table3}).
The aim of the present study was to evaluate the relationship between cardiopulmonary exercise testing, obstructive jaundice (especially severe obstructive jaundice with serum bilirubin $>$ 250 $\mu$mol/l) and preoperative pathophysiology in patients undergoing pancreaticoduodenectomy.

@@

\subsection{Patients}
Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and December 2012 and had undergone cardiopulmonary exercise testing as part of their preoperative assessment at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study. 
Established criteria for resectability in patients with malignant disease were used as outlined in Section \ref{sec:resectability_criteria}. 
Segmental or wedge resection of the portal vein or superior mesenteric vein was carried out if the lesion was otherwise resectable.

\subsection{Preoperative data}
Patient demographics, preoperative clinico-pathological characteristics including cardiorespiratory comorbidity, results of preoperative blood tests, chest x-ray, ECG and cardiopulmonary exercise tests were collected from prospectively held databases. 
The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease, respiratory disease, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) and was used as an objective score of comorbidity (Table \ref{table:intro_possum} on p\pageref{table:intro_possum}).
Cardiovascular comorbidity was defined as a score of 2 or more for either the cardiac disease or ECG component of the POSSUM score. 
Respiratory comorbidity was defined as a score or 2 or more for the respiratory disease component of the POSSUM score.

\subsection{Obstructive jaundice}
Serum bilirubin levels and liver function tests were measured in all patients on the day before surgery. 
Obstructive jaundice was defined as serum bilirubin $>$ 35 $\mu$mol/litre and severe obstructive jaundice was defined as serum bilirubin $>$ 250 $\mu$mol/litre. 
These thresholds were chosen as a recent randomised controlled trial that demonstrated increased complications in patients who underwent PBD excluded patients with serum bilirubin $>$ 250 $\mu$mol/litre from the study \parencite{van_der_gaag_preoperative_2010}.
The present study aimed to evaluate preoperative pathophysiology in this particular group of patients with `severe obstructive jaundice'.
Due to referral practices at the time of this study, a proportion of patients underwent preoperative biliary interventions and drainage before referral to the West of Scotland Pancreatic Unit. 
Data regarding biliary intervention and stenting was collected where feasible.

\subsection{Cardiopulmonary exercise test}
Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
All patients underwent standard pulmonary function tests and spirometry prior to cardiopulmonary exercise testing. 
A cycle ergometer was used to perform a symptom-limited, incremental work-load test preceded by a 3-minute rest period. 
The test was stopped when patients achieved their maximum exercise tolerance, when significant ischaemic changes occurred on ECG or for other safety reasons. 
Peak oxygen consumption achieved at this stage was defined as $\dot{V}_{O_2}$Peak. 
The $\dot{V}_{O_2}$AT was calculated using the V-slope \parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents \parencite{society_ats/accp_2003} methods.
$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us (Chapter \ref{ch_cpet_outcomes}) as well as Ausania and co-workers \parencite{ausania_effects_2012} which has shown increased incidence of complications in patients with $\dot{V}_{O_2}$AT below this threshold. 
Oxygen consumption at peak exercise ($\dot{V}_{O_2}$Peak) was dichotomised using a cut-off of 16 ml/kg/min. 
Cardiopulmonary exercise testing methodology and calculation of $\dot{V}_{O_2}$AT is described in detail in Sections \ref{sec:cpx_method} and \ref{sec:cpx_at_method}.
The cardiopulmonary exercise testing parameters described in this study are explained in detail in Section \ref{sec:cpx_parameters} on p\pageref{sec:cpx_parameters} and in published literature \parencite{balady_clinicians_2010}.

\subsection{Statistics}
Grouping of the variables was carried out using standard or previously published thresholds. 
In the absence of such thresholds, the variables were treated as continuous variables. 
Non-parametric tests were used to analyse the association between categorical and continuous variables while Chi-square tests were used to analyse the association between categorical variables. 
Univariate and multivariate binary logistic regression analysis was used to study the relationship between preoperative patient characteristics and $\dot{V}_{O_2}$AT / $\dot{V}_{O_2}$Peak. 
Scatter-plots were used for visual representation of the relationship between serum bilirubin and $\dot{V}_{O_2}$.

The level of significance was set at $p<0.05$.
SPSS software (Version 17.0; SPSS Inc., Chicago, IL, USA) was used to perform statistical analysis.

@@

The baseline demographic and clinical characteristics of non-jaundiced and jaundiced patients are shown in Table \ref{table:cpet_oj_patient}. 
A greater proportion of jaundiced patients were females (p=0.028) and smokers (p=0.038) compared to the non-jaundiced cohort. 
Elevated POSSUM Physiology Score (p=0.004) and malignancy (p$<$0.001) were significantly associated with the presence of jaundice. 
However, there was no statistically significant difference in age, body mass index, cardiovascular comorbidity, respiratory comorbidity or PBD between the non-jaundiced and jaundiced patients.

@@

\subsection{Obstructive jaundice vs. pulmonary function tests}
Pulmonary function tests including forced vital capacity (FVC), forced expiratory volume in one second (FEV1) and their derived parameters (predicted FEV1, FEV1/FVC, predicted FEV1/FVC) were compared between jaundiced and non-jaundiced patients. 
There was a trend towards lower forced vital capacity with increasing severity of jaundice but this did not reach statistical significance (p=0.092). 
There was no association between the other pulmonary function tests and jaundice (Table \ref{table:cpet_oj_pft}).

@@

\subsection{Obstructive jaundice vs. CPET}
Cardiopulmonary exercise test parameters measured at the anaerobic threshold and at peak exercise were compared between jaundiced and non-jaundiced patients using the Kruskal-Wallis test. 
Obstructive jaundice was associated with lower tidal volume (p=0.017), lower absolute $\dot{V}_{O_2}$ (p=0.003), lower corrected $\dot{V}_{O_2}$ (p=0.029), lower oxygen pulse (p=0.037) and lower respiratory rate (p=0.022) at the anaerobic threshold.
At peak exercise, jaundice was associated with lower tidal volume (p=0.028), lower absolute $\dot{V}_{O_2}$Peak (p=0.007), lower absolute $\dot{V}_{CO_2}$ (p=0.016) and lower end-tidal $O_2$ or $PET_{O_2}$ (p=0.026). 
However, as shown in Tables \ref{table:cpet_oj_anaerobic} and \ref{table:cpet_oj_peak}, several of these associations were not linear with a trend towards poorer results between non-jaundiced and mildly jaundiced patients and better values in the severely jaundiced cohort.

@@

The results of the present study demonstrate that obstructive jaundice, even when severe, is not an independent risk factor for poor preoperative cardiopulmonary exercise physiology in patients undergoing pancreaticoduodenectomy. 
The results also demonstrate that obstructive jaundice is associated with elevated preoperative systemic inflammation with the severity of systemic inflammation related to the severity of the jaundice.
The results also appear to suggest that there may be multiple other determinants of aerobic capacity including female sex, body mass index, haemoglobin, presence of cancer and systemic inflammation.

%History of CPET
The use of CPET in preoperative risk prediction was first made popular over two decades ago by Older and co-workers \parencite{older_preoperative_1993}. 
Since then cardiopulmonary exercise testing has been reported to be useful in identifying high risk patients prior to major general \parencite{snowden_submaximal_2010}, pancreatic \parencite{ausania_effects_2012}[Chapter \ref{ch_cpet_outcomes} of this thesis], oesophagogastric \parencite{nagamatsu_preoperative_2001} as well as vascular \parencite{carlisle_mid-term_2007} surgery. 
Cardiopulmonary exercise testing is superior to conventional measures of fitness due to the dynamic nature of the test that evaluates the adequacy of oxygen delivery to tissues under physiological stress. 
However, the factors responsible for poor aerobic capacity in surgical patients have not been adequately studied. 
The present study aimed to address this in patients undergoing major pancreatic surgery.

@@

Historically, obstructive jaundice has also been reported to be associated with adverse haemodynamic events in patients undergoing major surgery. 
Intra-operative blood loss, postoperative hypotension, increased susceptibility to shock and renal dysfunction were more common in patients with obstructive jaundice \parencite{dixon_factors_1983, pain_perioperative_1985, green_systemic_1995}.
A recent review noted that bile acids had a complex range of receptor mediated effects on the cardiovascular system \parencite{khurana_bile_2011}. 
Moreover, some of these effects appear to be partly reversible by biliary drainage as demonstrated by Padillo and co-workers \parencite{padillo_improved_2001}.

These observations led to routine pre-operative biliary drainage being recommended in all jaundiced patients before undertaking major surgery. 
Pancreaticoduodenectomy was described by Whipple initially as a two-stage operation, with the first stage involving a cholecysto-gastrostomy aimed at relieving biliary obstruction before undertaking the resection at a later second operation \parencite{whipple_treatment_1935}.

%Reasons for not doing PBD
However, several recent studies have reported that PBD is associated with increased incidence of complications and that surgery in the jaundiced patients is safe.
Pitt and co-workers in a prospective randomised trial compared outcomes in jaundiced patients undergoing surgery with or without PBD. They reported that PBD was associated with increased cost without any decrease in postoperative complications \parencite{pitt_does_1985}. 
A recent meta-analysis \parencite{sewnath_meta-analysis_2002} pooled data from 5 randomised controlled trials comparing surgery with PBD versus surgery without drainage.
The authors concluded that PBD not only did not improve postoperative complication rates or mortality but resulted in a higher overall complication rate due to the morbidity associated with the procedure itself. 
These findings have been reproduced in a more recent multicenter, randomised trial \parencite{van_der_gaag_preoperative_2010}.
A recent Cochrane Collaboration review of six trials including 520 patients concluded that PBD may be associated with serious adverse events and must not be performed routinely outside trial settings \parencite{wang_preoperative_2008}.

%CPET vs Jaundice
In the present study, obstructive jaundice was not independently related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak.
There was only a weak negative correlation between serum bilirubin and $\dot{V}_{O_2}$AT in our cohort.
These findings are similar to those of Parker and co-workers who reported that there was only a weak negative correlation between serum bilirubin and  $\dot{V}_{O_2}$Peak (Pearson's $\rho$ -0.21, p=0.02) and no correlation with $\dot{V}_{O_2}$AT ($\rho$ = -0.15, p=NS) in patients undergoing pancreaticoduodenectomy \parencite{parker_serum_2014}. 
Junejo and co-workers studied oxygen extraction in 9 jaundiced patients during cardiopulmonary exercise test by placing catheters within the femoral vein and reported that oxygen extraction was normal during cardiopulmonary exercise even in the presence of jaundice \parencite{junejo_peripheral_2014}

@@

%Obstructive jaundice and systemic inflammation
However, jaundice was associated with elevated preoperative systemic inflammation as measured by C-reactive protein levels and serum albumin. 
Obstructive jaundice is known to be associated with wide ranging abnormalities of both humoral and innate immune systems \parencite{nehez_compromise_2002, padillo_cytokines_2001, scott-conner_pathophysiology_1994}. 
It is unclear if this preoperative immune dysfunction continues after surgery as an abnormal postoperative systemic inflammatory response and this is explored further in Chapter \ref{ch_pre_post_sirs}.
However, studies suggest that the immunological abnormalities associated with obstructive jaundice did not improve after biliary drainage \parencite{kimmings_endotoxin_2000}.

@@

%\subsection{Conclusion}
In conclusion, obstructive jaundice, including severe obstructive jaundice did not affect preoperative cardiopulmonary exercise physiology. 
Therefore, preoperative biliary drainage cannot be expected to improve aerobic capacity in patients undergoing pancreaticoduodenectomy.
Both obstructive jaundice and low $\dot{V}_{O_2}$AT were associated with elevated preoperative systemic inflammation and provide further evidence of the role of jaundice and impaired aerobic capacity in immune dysregulation in these patients.

### 2015-08-26 Chapters/CPET_Outcomes.tex
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes

\section{Introduction}
Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 16-17\% surviving beyond the first year and 3\% surviving beyond 5 years \parencite{cancerresearchuk_cancer_2014}. 
The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2014, sener_pancreatic_1999}. 
In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000, geer_prognostic_1993} followed by adjuvant chemotherapy in selected patients \parencite{neoptolemos_randomized_2004,neoptolemos_adjuvant_2009} remains the primary modality of cure.

@@

%Figure 1
\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.8\linewidth]{Figures/cpet_outcomes_dist_of_AT}
	\caption{Distribution of $\dot{V}_{O_2}$AT across the study population.}
	\label{fig:cpet_outcomes_dist_of_AT}
\end{figure}

@@

% Figure 2
\begin{figure}[h]
\centering
\includegraphics[width=0.8\linewidth]{Figures/cpet_outcomes_km_at_los}
\includegraphics[width=1\linewidth]{Figures/cpet_outcomes_km_at_los_table} %Table below Kaplan Meier curve
\caption{Kaplan-Meier Plot of postoperative length of stay in patients with $\dot{V}_{O_2}$AT $>=$ 10 ml/kg/min versus $<$ 10 ml/kg/min.}
\label{fig:cpet_outcomes_km_at_los}
\end{figure}

@@

In the present study, $\dot{V}_{O_2}$AT was less than 10 ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. 
The proportion of patients with $\dot{V}_{O_2}$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit. 
While several studies have shown that low $\dot{V}_{O_2}$AT and/or low $\dot{V}_{O_2}$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this \parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010}. 
Older and co-workers reported in 1993 that low $\dot{V}_{O_2}$AT less than 11 ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.

However, Snowden and co-workers \parencite{snowden_submaximal_2010} reported that patients with an $\dot{V}_{O_2}$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers \parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the $\dot{V}_{O_2}$AT did not predict postoperative adverse events less after oesophagectomy. 
The lack of association between low $\dot{V}_{O_2}$AT and cardiopulmonary complications in this study may have been due to two reasons. 
Major cardiopulmonary complications occurred more often in association with major intra-abdominal adverse events which are determined largely by pancreatic morphology and local anatomy \parencite{braga_prognostic_2011}. 
Moreover, the stringent fitness criteria for undergoing pancreaticoduodenectomy may have excluded patients with known co-morbid cardiorespiratory diseases such as severe chronic obstructive pulmonary disease or cardiac failure.

>>>>

\section{Introduction}
Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 21\% surviving beyond the first year and 3\% surviving beyond 5 years \parencite{cancerresearchuk_cancer_2011}. 
The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2011, sener_pancreatic_1999}. 
In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000, geer_prognostic_1993} followed by adjuvant chemotherapy in selected patients \parencite{neoptolemos_randomized_2004,neoptolemos_adjuvant_2009} remains the primary modality of cure.

@@

%Figure 1
\begin{figure}[p]
	\centering
	\includegraphics[width=\textwidth]{Figures/cpet_outcomes_dist_of_AT}
	\caption{Distribution of $\dot{V}_{O_2}$AT across the study population.}
	\label{fig:cpet_outcomes_dist_of_AT}
\end{figure}

@@

% Figure 2
\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{Figures/cpet_outcomes_km_at_los}
\includegraphics[width=\textwidth]{Figures/cpet_outcomes_km_at_los_table} %Table below Kaplan Meier curve
\caption{Kaplan-Meier Plot of postoperative length of stay in patients with $\dot{V}_{O_2}$AT $>=$ 10 ml/kg/min versus $<$ 10 ml/kg/min.}
\label{fig:cpet_outcomes_km_at_los}
\end{figure}

@@

In the present study, $\dot{V}_{O_2}$AT was less than 10 ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. 
The proportion of patients with $\dot{V}_{O_2}$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit.

While several studies have shown that low $\dot{V}_{O_2}$AT and/or low $\dot{V}_{O_2}$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this \parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010}.

Older and co-workers reported in 1993 that low $\dot{V}_{O_2}$AT less than 11 ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
However, Snowden and co-workers \parencite{snowden_submaximal_2010} reported that patients with an $\dot{V}_{O_2}$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers \parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the $\dot{V}_{O_2}$AT did not predict postoperative adverse events less after oesophagectomy.

The lack of association between low $\dot{V}_{O_2}$AT and cardiopulmonary complications in this study may have been due to two reasons. 
Major cardiopulmonary complications occurred more often in association with major intra-abdominal adverse events which are determined largely by pancreatic morphology and local anatomy \parencite{braga_prognostic_2011}. 
Moreover, the stringent fitness criteria for undergoing pancreaticoduodenectomy may have excluded patients with known co-morbid cardiorespiratory diseases such as severe chronic obstructive pulmonary disease or cardiac failure.
